---
title: "SplitAlignerR Quick Start"
vignette: >
  %\VignetteIndexEntry{SplitAlignerR Quick Start}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

SplitAlignerR aligns empirical gene trees to the primitive branch-coordinate
axis of one reference tree. The production state logic is implemented once in
C++17. R converts inputs, calls that core, and wraps the result. The bundled
Catnip10 Oracle is deliberately separate pure R node-edge graph surgery; it is
benchmark evidence, not a second production implementation.

## The three result layers

Keep these questions separate:

1. `state_matrix`: what happened to each primitive reference coordinate after
   taxon restriction and empirical split recovery?
2. `numeric_matrix`: which primitive or composite coordinates have accepted
   finite branch-length evidence?
3. provenance: which frozen primitive members define each composite coordinate?

The single-tree states are `mapped`, `NA_struct`, `NA_fuse`, and `NA_topo`.
Paired `residual_NA` is the summary name for finalized literal `NA`; it is not
a fifth graph state and it is never written into a finalized matrix.

## Align one or more gene trees

```{r align}
library(SplitAlignerR)

species <- "((A:1,B:1):1,(C:1,D:1):1);"
genes <- c(
  concordant = "((A:1,B:1):1,(C:1,D:1):1);",
  discordant = "((A:1,C:1):1,(B:1,D:1):1);",
  marker = "((A:NaN,B:1):1,(C:1,D:1):1);"
)

result <- align_branches(species, genes, mode = "free")
result
result$state_matrix
result$numeric_matrix
```

The discordant tree lacks the reference internal split, so that primitive cell
is `NA_topo`. The `marker` tree still recovers the terminal split for `A`, so
its state is `mapped`; only its numeric value is unavailable. Software failure
markers are never converted to zero and never manufacture topological absence.

`species_tree` may also be one `ape::phylo` object or a file containing one
Newick tree. `gene_trees` may be a character vector, one `phylo`, a named list
of `phylo` objects, an `ape::multiPhylo`, or a line-based file. File records may
be plain Newick or use the bounded `gene_id(tree);` form.

## Optional paired bookkeeping

Fixed and free single-tree results remain independent objects. Pair them only
when their ordered gene and coordinate axes and conventions agree.

```{r paired}
fixed <- align_branches(species, genes["concordant"], mode = "fixed",
                         gene_ids = "g1")
free <- align_branches(species, genes["discordant"], mode = "free",
                        gene_ids = "g1")
paired <- pair_alignment_results(fixed, free)
paired$semantic_state_matrix
paired$free_pre_promotion_matrix
paired$final_matrix
paired$residual_NA
```

The semantic state matrix is the unchanged free single-tree ledger. The paired
ledger separately records the free pre-promotion token, fixed primitive and
fused numeric availability, final matrix token, and summary class. Only finite
fixed primitive evidence can promote generic free `NA` to `NA_topo`; a finite
fixed fused coordinate cannot satisfy that gate. Final literal `NA` is retained
as `NA` and summarized as `residual_NA`. Empirical split conflict or adjacent
side-clade loss may explain a frozen data set, but neither is a classification
predicate.

## Structural restriction and composite provenance

With only `A` and `B` retained, their primitive terminal coordinates form one
composite path while coordinates on the lost side become structural absence.

```{r provenance}
pruned <- align_branches(species, c(two_taxa = "(A:4,B:5);"))
pruned$state_matrix
pruned$composite_ledger
coordinate_provenance(pruned, "F[B1|B2]")
```

Numeric evidence for a fused path lives on the composite column in
`numeric_matrix`. It is not copied back into its primitive `NA_fuse` cells.
Composite coordinates are provenance-bearing member sets; this package does not
claim that they form an orthogonal, additive, or independent basis.

## Summaries and diagnostics

```{r summary}
summary(result)
subset(result$diagnostics, count > 0L)
```

Species-tree branch lengths, internal support/labels, annotation blocks, and a
representation-root length are parsed and ignored with structured diagnostics.
Gene-tree internal support/labels and annotation blocks are also ignored. Gene
tip labels, topology, and branch-length evidence are retained. Duplicate taxa,
unexpected taxa, invalid numeric tokens, parse failures, and incompatible
result schemas fail loudly.

Finite negative values are retained with a diagnostic because the nonnegative
theorem scope does not cover them. Decimal overflow and underflow to zero are
rejected. Conventional textual spellings such as `NaN` and `Inf` are recognized
before numeric conversion and classified as unavailable software evidence.

```{r numeric-policy}
validate_branch_length_tokens(c("0", "-0.1", "NaN", "Inf", "1e9999"))
```

## Save and reload without flattening layers

```{r persistence}
path <- tempfile(fileext = ".rds")
save_splitaligner_result(result, path)
restored <- read_splitaligner_result(path)
identical(result, restored)
```

Reload checks the stored core and result-schema versions by default. Existing
files are not overwritten unless `overwrite = TRUE` is explicit.

## Benchmark track

```{r benchmark}
validate_catnip10_oracle()
rebuilt <- recompute_catnip10_oracle("global")
identical(rebuilt$matrix, catnip10_matrix("global"))
```

The Catnip10 benchmark is discordance-free. Its two cumulative deletion regimes
contain 272 frozen primitive cells in total. The pure R Oracle recomputes their
graph states independently of the C++ mapper. Development tests additionally
cross-check the mapper against every cell and frozen fusion member set. These
are implementation gates, not final release certification.
